Relict deciduous oaks show high resilience to drought despite the effect of land-use legacies

### 2019-04-22 man/ms.Rmd
update

* The studied relict populations showed expressions of management, likely reflecting the origin of the current forest structure. Both resilience indices of primary and secondary growth expressed that trees showed high resilience both to recent drought events and also to long-term climatic changes (i.e. warming), despite the ‘a priori’ vulnerability that we were expecting them to express in respone climate change on relict stands resulting from intensive land-use in history at their current rear-edge. Furthermore, the coldest stands show a positive growth trend since the end of the 1970s matching the simultaneous increase in temperature observed.

@@

The values of these indices were computed for tree growth (BAI) and greenness (EVI mean) during each drought event. 2005 and 2012 were considered as single drought events. The predrought and postdrought values of each target variable (*i.e.*: BAI or EVI) were computed as the mean value during a period of three years before and after the drought event respectively. A period of three years was chosen because we found similar results comparing periods of two, three and four years (Figure S6) and this length was used in other studies [*e.g.* @Gazol2017]. To put in context in a longer-term the resilience observed for 2005 and 2012 drought events, as well as to test whether the resilience indices were ecologically meaningful, resilience metrics for BAI data were additionally computed for the most severe drought events since 1900 and compared them with drought severity (Figure 6, Table S3).

@@

During the last two drought events, resilience metrics for greenness and tree-growth were significantly different bewteen drought events (Table 3). We obtained significantly higher Resilience (*Rs*) and Resistance (*Rt*) values for the 2012, the most severe event, than for 2005 in both variables (Tables S1-S2; Figure 7). Recovery (*Rc*) showed contrasting pattern for EVI and tree-growth. Recovery values for greenness was higher for 2005 than for 2012 drought event (Table S1, Figure 7).

@@

Our findings show that severe droughts, such us 2005 and 2012, provoked a reduction both in greenness (*i.e.* primary growth) and in secondary growth of *Q. pyrenaica* (Figures 2 and 4). These results are consistent with previous studies that reported significant reductions of tree-growth for this oak species during extreme drought events [@Corcuera2006; @GeaIzquierdo2014; @RubioCuadrado2018]. The greatest reduction of tree-growth was observed during the 1995 drought, a characteristic pointer-year that caused severe and extensive damage to the Mediterranean vegetation across Iberian Peninsula [@Penuelas2001; @Camarero2018; @Gazol2018].

During the last decades, the south of Iberian Peninsula suffered two of the worst recorded drought [2005 and 2012; @GarciaHerrera2007; @Trigo2013; @VicenteSerrano2014] (Table S3). However, we found a positive trend for vegetation greenness of *Q. pyrenaica* (78.9 % of the pixels showed positive trend  during the 2000 to 2016 period). This confirms previous findings that pointed out an increase in primary productivity for these forests [@PerezLuque2015onto], being the most productive ecosystems in this mountain region during that period [@Alcaraz2016obsnev_ndvi]. For tree-growth, we also observed positive trends in the last decades, particularly for the southern high-elevation site (CA-High, Figure 4). Similar long-term trends were described for this species along their distribution range [@GeaIzquierdo2014; @RubioCuadrado2018] and could be related to the rising of the temperatures in the last decades [@GeaIzquierdo2014]. Our pattern differs from the decline trends observed for several mediterranean and temperate tree-species located in their rear-edges [*e.g.*, *Pinus nigra*; *P. sylvestris*; @SanchezSalguero2012; @Camarero2015; *Fagus sylvatica*; @Dorado2017].

Resilience (*Rs*) values observed in our study sites for the latest drought events in both tree-growth and greenness were close to 1, except for the tree-growth during 2005 event (Tables S1 and S2). Despite the 2012 drought event was more severe and intense than 2005, according with our analysis with SPEI data (Table S3), resilience values for greenness and tree-growth were greater for 2012 than for 2005. These results could be explained by the different timing of the two droughts, *i.e.* the 2012 drought was a winter drought [@Trigo2013] that occurred earlier than the 2005 drought. The latter lasted less than 2012 drought, but matched the period of maximum growth for *Q. pyrenaica* forests [@PeresLis2017]. Our results are in line with recently studies that indicate the timing of the drought as a key factor determining tree recovery after drought [@Huang2018],  that also has been observed for other Mediterranean oak species [*e.g.* Holm oak; @Camarero2015b].

@@

Rear-edge populations live in environmental narrow margins, and small variations in environmental conditions can increase the vulnerability of the species [@Hampe2005]. For instance, higher vulnerability to drought is ofted assumed for rear-edge populations [@MartinezVilalta2018], however some studies showed different findings [@Cavin2017; @Granda2017], as our results suggest, with high values of resilience for populations of * Q. pyrenaica * located in its rear-edge.

This would be related with the definition of the marginal population. Algunos autores han apuntando que cuando se estudian poblaciones del rear-edge, hay que poner atención a la forma en la que se define la marginalidad, esto es, si se define atendiendo a criterios geográficos, climáticos, o según otros factores ecológicos [@MartinezVilalta2018]. En este sentido, los altos valores de resiliencia a los eventos de sequía que hemos observado, podrían sugerir que las poblaciones de *Q. pyrenaica* en Sierra Nevada están situados en un rear-edge geográfico, pero no climático. Esto último se ve reforzado por el hecho de que Sierra Nevada is considered a glacial refugia for deciduous *Quercus* species [@Brewer2002; @Olalde2002; @RodriguezSanchez2010]; además las poblaciones de *Q. pyrenaica* en Sierra Nevada presentan una alta resiliencia genética [@Valbuena2013; @Valbuena2017]

Several considerations have to be taken into account to explain this result. Firstly it would be interesting to explore the cumulative precipitation of the previous and posterior years of these drought events. For instance, the 1992-1994 years (pre-1995 drought) had precipitation values below the mean of the whole period (Figure S1), while 1996-1998 years (post-1995 drought) showed a wet opposite pattern. In fact, the first months of 1996 recorded the highest cumulative rainfall of the past century for our southern site (Cáñar) [@MesaTorres2009]. In addition, this argument could also explain the higher values of recovery found for 1995 (Figure 6).

>>>>

* The studied relict populations showed expressions of management, likely reflecting the origin of the current forest structure. Both resilience indices of primary and secondary growth expressed that trees showed high resilience both to recent drought events and also to long-term climatic changes (*i.e.* warming), despite the ‘a priori’ vulnerability that we were expecting them to express in respone climate change on relict stands resulting from intensive land-use in history at their current rear-edge. Furthermore, the coldest stands show a positive growth trend since the end of the 1970s matching the simultaneous increase in temperature observed.

@@

The values of these indices were computed for tree growth (BAI) and greenness (EVI mean) during each drought event. 2005 and 2012 were considered as single drought events. The predrought and postdrought values of each target variable (*i.e.* BAI or EVI) were computed as the mean value during a period of three years before and after the drought event respectively. A period of three years was chosen because we found similar results comparing periods of two, three and four years (Figure S6) and this length was used in other studies [*e.g.* @Gazol2017]. To put in context in a longer-term the resilience observed for 2005 and 2012 drought events, as well as to test whether the resilience indices were ecologically meaningful, resilience metrics for BAI data were additionally computed for the most severe drought events since 1900 and compared them with drought severity (Figure 6, Table S3).

@@

During the last two drought events, resilience metrics for greenness and tree-growth were significantly different bewteen drought events (Table 3). The 2005 drought event reduced greennes and growth more than that of 2012 (Tables S1 and S2). We obtained significantly higher Resilience (*Rs*) and Resistance (*Rt*) values for the 2012, the most severe event, than for 2005 in both variables (Tables S1-S2; Figure 7). Recovery (*Rc*) showed contrasting pattern for EVI and tree-growth. Recovery values for greenness was higher for 2005 than for 2012 drought event (Table S1, Figure 7).

@@

Severe drought events negatively affected to primary and secondary growth of *Q. pyrenaica* forests in the rear edge of their distribution. Our findings show that 2005 and 2012 drought events, provoked reductions in greenness and tree-growth of these forests. These results are consistent with previous studies that reported significant reductions of tree-growth for this oak species during extreme drought events [@Corcuera2006; @GeaIzquierdo2014; @RubioCuadrado2018]. In fact, when we explore the effects of drought at a longer-term scale, we observed the greatest reduction of tree-growth during the 1995 drought, a characteristic pointer-year that caused severe and extensive damage to the Mediterranean vegetation across the Iberian Peninsula [@Penuelas2001; @Camarero2018; @Gazol2018].

Although the south of Iberian Peninsula suffered severe drought events in the last decades [*e.g* 2005 and 2012 were considered two of the worst recorded drought; @GarciaHerrera2007; @Trigo2013; @VicenteSerrano2014] (Table S3), we found a positive trend for vegetation greenness of *Q. pyrenaica* (78.9 % of the pixels showed a positive trend during the 2000 to 2016 period). This confirms previous findings that pointed out an increase in primary productivity for these forests [@PerezLuque2015onto], being the most productive ecosystems in this mountain region during that period [@Alcaraz2016obsnev_ndvi]. For tree-growth, we also observed positive trends in the last decades, particularly for the southern high-elevation site (CA-High, Figure 4). Similar long-term trends were described for this species along their distribution range [@GeaIzquierdo2014; @RubioCuadrado2018] and could be related to the rising of the temperatures in the last decades [@GeaIzquierdo2014]. Howewer, the observed pattern for our sites (a rear-edge) differs from the decline trends observed for several Mediterranean and temperate tree-species located in their rear-edges [*e.g.*, *Pinus nigra*; *P. sylvestris*; @SanchezSalguero2012; @Camarero2015; *Fagus sylvatica*; @Dorado2017].

Resilience (*Rs*) values observed in our study sites for the latest drought events in both tree-growth and greenness were close to 1, except for the tree-growth during 2005 event (Tables S1 and S2). Despite the 2012 drought event was more severe and intense than 2005, according with our analysis with SPEI data (Table S3), resilience values for greenness and tree-growth were greater for 2012 than for 2005. These results could be explained by the different timing of the two droughts, *i.e.* the 2012 drought was a winter drought [@Trigo2013] that occurred earlier than the 2005 drought. The latter lasted less than 2012 drought, but matched the period of maximum growth for *Q. pyrenaica* forests [@PeresLis2017] (Figure S7). Our results are in line with recently studies that indicate the timing of the drought as a key factor determining tree recovery after drought [@Huang2018],  that also has been observed for other Mediterranean oak species [*e.g.* Holm oak; @Camarero2015b].

@@

Rear-edge populations live in environmental narrow margins, and small variations in environmental conditions can increase the vulnerability of the species [@Hampe2005]. For instance, higher vulnerability to drought is often assumed for rear-edge populations [@MartinezVilalta2018], however some studies showed different findings [@Cavin2017; @Granda2017], as our results suggest, with high values of resilience for populations of *Q. pyrenaica* located in its rear-edge. This would be related with the definition of the marginal population.
Algunos autores han apuntando que cuando se estudian poblaciones del rear-edge, hay que poner atención a la forma en la que se define la marginalidad, esto es, si se define atendiendo a criterios geográficos, climáticos, o según otros factores ecológicos [@MartinezVilalta2018]. En este sentido, los altos valores de resiliencia a los eventos de sequía que hemos observado, podrían sugerir que las poblaciones de *Q. pyrenaica* en Sierra Nevada están situados en un rear-edge geográfico, pero no climático. Esto último se ve reforzado por el hecho de que Sierra Nevada is considered a glacial refugia for deciduous *Quercus* species [@Brewer2002; @Olalde2002; @RodriguezSanchez2010]; además las poblaciones de *Q. pyrenaica* en Sierra Nevada presentan una alta resiliencia genética [@Valbuena2013; @Valbuena2017]

Several considerations have to be taken into account to explain the high resilience values found. Firstly it would be interesting to explore the cumulative precipitation of the previous and posterior years of these drought events. For instance, the 1992-1994 years (pre-1995 drought) had precipitation values below the mean of the whole period (Figure S1), while 1996-1998 years (post-1995 drought) showed a wet opposite pattern. In fact, the first months of 1996 recorded the highest cumulative rainfall of the past century for our southern site (Cáñar) [@MesaTorres2009]. This could explain the higher values of recovery found for 1995 (Figure 6).

VAS POR AQUI

De forma general: 
* dependencia del agua
* sitaución en Sierra Nevada
* Mayor resiliencia (mayor crecimiento) los robles mas altos

* Existencia de aportes extra de humedad. Las acequias

- Por un lado una alta disponibilidad de agua. Conocemos que esta especies es mas sensitive a la humedad del suelo que otras especies arbóreas mediterráneas (GeaFEM)

Por un lado la dependencia del agua. Sabemos que esta especie es mas sensitive a la humedad del suelo que otras especies arbóreas mediterráneas (GeaFEM). En Sie
